---
title: "Course presentation: statistical models link data to theory"
date: "2024-01-29" 
author:
  - name: Sara Mortara & Diogo Melo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Slides

```{r slides}
#| eval: true
xaringanExtra::embed_xaringan("01_presentation/slides.html")
```

# Further reading

 1. **Quantitative Reasoning in Ecology**: Chapter 1 in Schneider,
    D. 2009. Quantitative Ecology: Measurement, Models, and
    Scaling. New York, Academic Press, 2nd Ed. 
	*(Nice synthesis of
    statistical analyses in the context of quantitative reasoning. It has
    also many important highlights on quantities and measurements)*
 2. **Alternative views of scientific method and of modelling**:
    Chapter 2 in Hilborn, R & Mangel, M. 1997. The Ecological
    Detective - Confronting Models with Data. Princeton, Princeton
    University Press. 
	*(The whole book is a [true classic](https://books.google.com.br/books?id=pQabBQAAQBAJ&pg=PA3&hl=pt-BR&source=gbs_toc_r&cad=3#v=onepage&q&f=false) on
    the use of statistical models in ecology. This chapter has a quick
    epistemological review and highlights the key distinction between
    models and hypothesis. It also brings a typology of models, and
    the concept of model complexity.)*

