---
title: "Probability distributions"
description: Our first statistical models
author:
  - name: Paulo Inácio Prado
date: "2024-01-29"
output: 
    distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Slides

```{r slides}
#| eval: true
xaringanExtra::embed_xaringan("02_probability_distributions/slides.html")
```

# Lab

[Lab]()


# More learning resources

 - [Probability distribution
  viewer](https://github.com/ksmzn/ProbabilityDistributionsViewer), a
  interactive app and web page to play with the parameters of
  probability distributions. Check also [this
  video in Portuguese](https://youtu.be/MYwJNCYJAg0) about this resource.

 - Video lectures with more details on
 probability distributions, from a [more extensive Course on
 Statistical Models](http://insilvaarbores.com.br/BIE5781/)
 of the Graduate Program of Ecology at USP (in Portuguese):
   - [Introduction to probability distributions](https://youtu.be/5U72x-u-dvE)
   - [Discrete probability distributions](https://youtu.be/j9nCwfDWqkw)
   - [Introduction to continuous distributions](https://youtu.be/j9nCwfDWqkw)
   - [Continuous distributions for time](https://youtu.be/XsXjyPtTlk8)
   - [Normal, log-normal and uniform distributions](https://youtu.be/bO2e-y9E5mg)
 
 - [Python notebooks](https://mybinder.org/v2/gh/piLaboratory/bie5782/master?filepath=jupyter%2F) with mathematical deductions of some simple MLEs (in portuguese).




