---
title: "Multiple regression and model comparison"
description: |
  Multiple regression and  model selection
author:
  - name: Paulo Inácio Prado (PI)
    url: 
    affiliation: 
    affiliation_url:
date: "2024-01-31"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Slides

```{r slides}
#| eval: true
xaringanExtra::embed_xaringan("04_multiple_regression/slides.html")
```

# Lab

[Multiple linear regression Lab](04_multiple_regression/lab.html)


# Further reading

+ Crawley. M.J. 2015. Statistics: an introduction using R. John Wiley & Sons Ltd.

+ Gotelli, N.J., Ellison, A.M., Ellison, S.E. and S.R.F.H.F.A.M., 2004. A Primer of Ecological Statistics. Sinauer Associates Publishers.

+ James, G., Witten, D., Hastie, T., Tibshirani, R., 2013. An Introduction to Statistical Learning: with Applications in R, 1st ed. 2013. Corr. 4th printing 2014 edition. ed. Springer, New York, NY.

+ McElreath, R., 2018. Statistical Rethinking: A Bayesian Course with Examples in R and Stan, 2nd ed. CRC Press.






