---
title: "Maximum likelihood method"
description: |
  What's the likelihood function, and how it can be used to fit statistical models to data.
author:
  - name: Paulo Inácio K. L. Prado (PI)
    url: 
    affiliation: 
    affiliation_url:
date: "2024-02-02"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Slides

```{r slides}
#| eval: true
#xaringanExtra::embed_xaringan("04_maximum_likelihood/slides.html")
```

```{r slides-2}
#| eval: true
#xaringanExtra::embed_xaringan("04_maximum_likelihood/slides_2.html")
```

# Lab

[Maximum likelihood fitting and inference]()

# Further reading

 * Likelihood and all that. chapter 6 in: Bolker, B.M. 2008
    Ecological Models and Data in R Princeton: Princeton University
    Press.
 * Hobbs, N.T. & Hilborn, R. 2006. Alternatives to statistical
   hypothesis testing in ecology: A guide to self-teaching. Ecological
   Applications: 16(1): 5-19.
 * The  Concept of Likelihood , chapter 2 *in*: Edwards, A.W.F., 1992
    Likelihood. Baltimore: John Hopkins University Press.   
 * Introduction, and Information and likelihood theory. chapters 1-2
  **in** Burnham, K. P., & Anderson, D. R. (2002). Model Selection and
  Multimodel Inference: A Practical-Theoretic Approach, 2nd ed. New
  York, Springer-Verlag.	
 * [Vignette of the package `bbmle`](https://cran.r-project.org/web/packages/bbmle/vignettes/mle2.pdf) 







