---
title: "Mixed-effects linear models"
description: |
  Linear models that mix random and fixed effects.
author:
  - name: Diogo Melo
    url: 
    affiliation: 
    affiliation_url: 
date: "2024-02-09"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Slides

```{r slides}
#| eval: true
xaringanExtra::embed_xaringan("10_mixed_models/slides/index.html")
```

[Open slides in new window](10_mixed_models/slides/index.html)

[Slides PDF](10_mixed_models/10_MixedModels.pdf)

# Lab

[Hierachical Models Lab](10_mixed_models/lab.html)

# Further reading

* Bates, D., Mächler M., Bolker, B.M. and Walker, S.C.. 2015. Fitting Linear Mixed-Effects Models Using lme4. Journal of Statistical Software 67 (1): 1-48.

* Bolker, B. M. et al. 2009. Generalized linear mixed models: a practical guide for ecology and evolution. Trends in ecology & evolution, 24(3): 127-135.

* The Generalized Linear Mixed Models FAQ page: https://bbolker.github.io/mixedmodels-misc/glmmFAQ

* Gelman, A., Hwang, J., & Vehtari, A. (2014). Understanding predictive information criteria for Bayesian models. Statistics and computing, 24(6), 997-1016.

* Gelman, Andrew, and Jennifer Hill. 2006. Data Analysis Using Regression and Multilevel/Hierarchical Models. Cambridge, England: Cambridge University Press.

* Zuur, A et al. 2009. Mixed Effects Models and Extensions in Ecology with R. Springer.




