---
title: "Dynamic models"
description: |
  Bridging dynamical and statistical models
author:
  - name: Paula Lemos-Costa and Diogo Melo
    affiliation: The University of Chicago and Princeton University
date: "2023-02-02"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Slides

```{r slides}
#| eval: true
xaringanExtra::embed_xaringan("09_dynamics_models/DynamicalModels.html")
```

```{r slides_paulinha}
#| eval: true
xaringanExtra::embed_xaringan("09_dynamics_models/slides.html")
```

# Lab


[Dynamical Models Lab](09_dynamic_models/lab.html)


# Further reading

## Theoretical developments
* ["Predicting coexistence in experimental ecological communities."](https://www.nature.com/articles/s41559-019-1059-z)
 Nature ecology & evolution 4.1 (2020): 91-100. Maynard, Daniel S., 
 Zachary R. Miller, and Stefano Allesina.

* ["Modelling ecological communities when composition is manipulated 
 experimentally."](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.14028)
 Methods in Ecology and Evolution (2022) Skwara, Abigail, Paula Lemos‐Costa, 
 Zachary R. Miller, and Stefano Allesina.

## Experimental studies
* ["Phytoplankton diversity affects biomass and energy production differently during community development."](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2435.13955) 
Ghedini, Giulia, Dustin J. Marshall, and Michel Loreau. Functional Ecology 36.2 (2022): 446-457.

