---
title: "Statistical models: linking data to theory"
description: |
  Serrapilheira Training Program in Quantitative Biology and Ecology - 2024
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Welcome to the Statistical Inference module of the
Serrapilheira Training Program in Quantitative Biology and
Ecology 2024. Statistical models are a useful tool to bridge theory
and data. Why do ecologists fit statistical models? There are three
main reasons: explore data, make inferences, and derive predictions.

## Goal

Our general goal is to provide appropriate conceptual foundations on
statistical thinking in ecology. We will teach you basic tools needed to connect
theory and hypothesis to data and show you additional paths you can
take to further explore more advanced topics. There are two main steps
in this endeavor: (i) the construction and fit of statistical models;
and (ii) the inferences that can be made with and about the models.

## Instructors

* [Sara Mortara]() re.green, [¡liibre!](https://liibre.netlify.app/)
* [Diogo Melo](https://damelo.net/), Princeton University
* [Paulo Inácio Prado](http://ecologia.ib.usp.br/let/doku.php?id=engl:prado:start), Universidade de São Paulo

### Instuctors from previous courses

* [Andrea Sánchez-Tapia](https://andreasancheztapia.netlify.app/), Universidade Federal do ABC, [¡liibre!](https://liibre.netlify.app/)
* [Paula Lemos-Costa](https://lemoscosta.weebly.com/), The University of Chicago


## Preparing for the course

Before the course starts, follow [these
instructions](00_instructions.html) to install R <!-- RStudio --> and
the necessary R packages in your computer, and also to recap the
basics on R we'll need to start.

## Course schedule 

```{r, echo = F}
schedule <- readr::read_csv("schedule.csv")
kableExtra::kable(schedule)
```

