---
title: "Mixed-effects and hierarchical models lab"
author:
 - "Paulo Inácio Prado"
 - "Diogo Melo"
date: "2023-01-31"
output:
    distill::distill_article:
      toc: true
      toc_depth: 3
---

```{r setup, echo=FALSE}
library(rmarkdown)
knitr::opts_chunk$set(eval = FALSE, fig.asp = 1)
# clipboard
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "Copy code <i class=\"fa fa-clipboard\"></i>",
    success_text = "Copied! <i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "Not copied 😕 <i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
  )
```


-<!-- 
Tadpoles rethinkg: proporcoes de sobrevivencia, sem random intercept e com random intercept: shrinkage de estimativas pooled e non-pooled. Usar lme4:glmer . Ver graficos shrinkage do Rethinking
desafio : alguma interpretação do efeito do tamanho do tanque sobre o shrikage

https://blogs.rstudio.com/ai/posts/2019-05-06-tadpoles-on-tensorflow/
https://github.com/waldronlab/Statistical-Rethinking/blob/master/2018-05-23%20Multilevel%20Models.Rpres=


Especie-area de varios arquipélagos: modelo logS x logA com slopes variados, com e sem efeito fixo do tipo de arquipélago. começar mostrando as regressões para cada arquipélago. Usar quap

Notação: ver slides Ogro

The commands below uses [tidyverse](https://www.tidyverse.org/) syntax
to select only data for insular systems^[that is, excludes data for
"islands" thata are actually habitat patches], and archipelagoes with
at least five islands. After that, we calculate the logarithms of the
area and number of species of each island, and then standardize the
logarithm of areas^[by subtracting the mean of log area from each
value and then by dividing by the standard deviation of the value of
log area], and center the logarithm of number of species^[by
subtracting the mean of log species richness from each value]. The
code also creates a numeric index for each study site (archipelago).

-->


# Introduction


We will use a data set with the number of species of ants recorded in
islands in different archipelagoes. Our goal is to fit linear models
to the relationship of the number of species as a function of island
area, in log scale.


Start loading the required libraries and the data (which is
available in the supplementary material in Ohyama *et al.* 2021):


```{r}
## load required libraries
library(tidyverse)
library(rethinking)
library(lme4)
library(bbmle)
## all data sets compiled by Ohyama et al
raw <- read.csv("https://github.com/piLaboratory/bie5781/raw/master/data/jbi14149-sup-0003-appendixs3.csv")
```

The commands below use [tidyverse](https://www.tidyverse.org/) syntax
to select only data for oceanic islands. After that, we calculate the
logarithms of the area and number of species of each island, and then
standardize the logarithm of areas^[by subtracting the mean of log
area from each value and then by dividing by the standard deviation of
the value of log area], and center the logarithm of number of
species^[by subtracting the mean of log species richness from each
value]. The code also creates a numeric index for each study site
(archipelago).

```{r , eval=TRUE}
## Data set for oceanic islands
## and transformed area and richness
islands <-
    filter(raw, Island.Type== "Oceanic") %>%
    mutate(l_area = log(Island_area),
           l_S = log(Species.Richness),
           ls_area = (l_area - mean(l_area))/sd(l_area),
           lc_S = l_S - mean(l_S),
           site_fac= factor(Study.location),
           site = as.integer(site_fac),
           site_id = factor(site))
```

Tidyverse provides ggplot2 package, that allows to make a quick plot
of the species-area relationship at each study site:


```{r }
p2 <-
    islands %>%
    ggplot(aes(x = ls_area, y = lc_S)) +
    geom_point() +
    facet_wrap(~ site_fac) +
    xlab("Standardized ln Island Area (m2)") +
    ylab("Centered ln Species Richness") +
    theme_bw()

p2
```

In the following sections we will fit and compare three alternative
models to this data, assuming that the response follows a Gaussian
distribution:

**Single slope and intercept:** a simple linear regression, that
estimates a single slope and a single intercept for the expected
linear relationship between the response and predictor:

$$
\begin{aligned} 
  y_i &\sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i &= \alpha + \beta x_i
\end{aligned} 
$$

Where $y_i$ is the centered value of the logarithm of species
richness, and $x_i$ is the standardized value of the logarithm of
island area, for each island $i$

**Independent slopes and intercepts for each site:** this model
corresponds to fit a separate linear regression for each archipelago
(study site). Thus, there is a independent intercept and slope for the
linear predictor at each site:

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i &= \alpha_{\text{site[i]}} + \beta_{\text{site[i]}} \, x_i
\end{aligned}
$$

Where $\alpha_{\text{site[i]}}$ and $\beta_{\text{site[i]}}$ are the
intercept and slope for the site of the observation $i$.

**Pooled slopes and intercepts across sites:** there is a different
slope and intercept for the linear predictor for each site, but pooled
in a assumed distribution. This is an hierarchical model, where slopes
and intercepts for each site follow each one some theoretical
distribution, usually a Gaussian distribution.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i &= \alpha_{\text{site[i]}} + \beta_{\text{site[i]}} \, x_i \\
  \alpha_{\text{site}} &\sim \text{Normal}(\mu_\alpha, \sigma_\alpha) \\
  \beta_{\text{site}} &\sim \text{Normal}(\mu_\beta, \sigma_\beta) \\
\end{aligned}
$$

By pooling the parameters in the same distribution we assume that
there is some dependence among them.

In the next sections we will show how to fit and compare these three
models using the Bayesian and Maximum Likelihood approaches.

## The Bayesian way


### Single slope and intercept

We need the values of the response and predictor variable for
each island, as shown in the first lines of the data set below:

```{r }
head(islands[, c("ls_area", "lc_S")])
```

We need also to define prior distributions for the three parameters of the
model. Here we use a log-normal prior for the slope, as we know that
the species-area relationship is positive. The chosen log-normal prior assigns a
low probability (p<0.05) to slopes larger than 1.5. The priors for the
intercept and standard deviation are less informative:

$$
\begin{aligned} 
  y_i & \sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i &= \alpha + \beta x_i \\
  \alpha & \sim \text{Normal}(\mu = 0, \sigma = 2)\\
  \beta & \sim \text{Log-normal}(\mu = 0, \sigma = 0.7)\\
  \sigma & \sim \text{Exponential}(\lambda = 1)
\end{aligned} 
$$

And here is the code to fit this model using the *rethinking* package:

```{r }
m0 <- ulam(
    alist(
        lc_S ~ dnorm( mu, sigma ) ,
        mu <- alpha + beta * ls_area,
        alpha ~ dnorm(0, 2),
        beta ~ dlnorm(0, 0.7), 
        sigma ~ dexp(1)
    ) , data = islands[, c("ls_area", "lc_S")] , chains=4 , cores=4 , log_lik=TRUE )
```

You get a summary of the posterior distributions of the parameters with:


```{r }
(m0.cf <- precis(m0, digits =3))
```

And we can use the mean of the posterior distributions of the intercept
and slope to draw the regression line on the data for each site:


```{r }
p2 +
    geom_abline(aes(intercept = m0.cf["alpha", 1], slope = m0.cf["beta",1]))
```

Assuming the same regression line for every site is a huge
underfitting.

### Different slopes and intercepts for each site

Clearly we need a model with more parameters to describe the
variations of the species-area relationship we see across study
sites. Thus, we can try to allow different intercepts and slopes for
each of the sites. Now we will need a third variable, a numeric index
that codes the site:

```{r }
head(islands[, c("ls_area", "lc_S", "site")])
tail(islands[, c("ls_area", "lc_S", "site")])
```

And we will call $\alpha_{\text{site[i]}}$ the intercept for each
observation $i$ for each site. The same applies to the slope
$\beta_{\text{site[i]}}$. As we have `r max(islands$site)` sites, we
have to specify `r max(islands$site)` priors for each intercept and 
`r max(islands$site)` priors for each slope. Using the notation
$\theta_{\text{site}} \sim \text{distribution()}$ as shorthand
for this bunch of priors, we can write the model as:

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i &= \alpha_{\text{site[i]}} + \beta_{\text{site[i]}} \, x_i \\
  \alpha_{\text{site}} & \sim \text{Normal}(\mu = 0, \sigma = 2)\\
  \beta_{\text{site}} & \sim \text{Log-normal}(\mu = 0, \sigma = 0.7)\\
  \sigma & \sim \text{Exponential}(\lambda = 1)
\end{aligned}
$$

For all intercepts and slopes we choose the same priors used for the
single intercept and slope of the previous model. We also kept the
same prior for single parameter $\sigma$ .

We can fit this model with:


```{r }
m1 <- ulam(
    alist(
        lc_S ~ dnorm( mu, sigma ) ,
        mu <- a[site] + b[site] * ls_area,
        a[site] ~ dnorm(0, 2),
        b[site] ~ dlnorm(0, 0.7),
        sigma ~ dexp(1)
    ) , data = islands[, c("ls_area", "lc_S", "site")] , chains=4 , cores=4 , log_lik=TRUE )
```

To show the summary of the posteriors of all parameters use the
argument *depth = 2* of function *precis*:

```{r }
(m1.cf <- precis(m1, digits =3, depth =2))
```

The code below arranges the posterior means of the intercept of slope
for each site as columns of a data frame, and adds the site name for
each one. We can then use this data frame to add the regression lines
 to the scatter plots for each site:
 
 
```{r }
nsites <- max(islands$site)
## Data frame with linear parameters and site name
m1.ab <- data.frame(
    site = 1:nsites,
    site_fac = levels(islands$site_fac),
    intercept = m1.cf[1:nsites,1],
    slope = m1.cf[(nsites+1):(2*nsites),1])
## Scatter plots + posterior regression lines
p2 +
    geom_abline(data = m1.ab,
                aes(intercept = intercept, slope = slope),
                col="navy")
```

As we would expect, the fit improved very much, but at the cost of
adding `r nrow(m1.cf) - nrow(m0.cf)` parameters to the model. We
switched from underfitting to a possible overfitting. 

### Pooled intercepts and slopes

A  possible middle-ground is to
allow varying parameters across sites, but pooling them in a
distribution. Thus, now we have **hyperparameters**, which are the
parameters of the distributions of the parameters of the linear
predictor. We thus have to assign priors for the hyperparameters:


$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i &= \alpha_{\text{site[i]}} + \beta_{\text{site[i]}} \, x_i \\
  \alpha_{\text{site}} & \sim \text{Normal}(\mu_\alpha, \sigma_\alpha)\\
  \beta_{\text{site}} & \sim \text{Normal}(\mu_\beta, \sigma_\beta)\\
  \mu_\alpha & \sim \text{Normal}(\mu = 0, \sigma = 2)\\
  \mu_\beta & \sim \text{Log-normal}(\mu = 0, \sigma = 0.7)\\
  \sigma_\alpha & \sim \text{Exponential}(\lambda = 1)\\
  \sigma_\beta & \sim \text{Exponential}(\lambda = 0.25)\\
  \sigma & \sim \text{Exponential}(\lambda = 1)\\
\end{aligned}
$$

Here is the code to fit this model:


```{r }
m2 <- ulam(
    alist(
        lc_S ~ dnorm( mu, sigma ) ,
        mu <- a[site] + b[site] * ls_area,
        a[site] ~ dnorm(mu_a, sigma_a),
        b[site] ~ dnorm(mu_b, sigma_b),
        ##prior
        sigma ~ dexp(1),
        ## hyper-priors
        mu_a ~ dnorm(0, 2),
        mu_b~ dlnorm(0, 0.7),
        sigma_a ~ dexp(1),
        sigma_b ~ dexp(0.25)
    ) , data = islands[, c("ls_area", "lc_S", "site")] , chains=4 , cores=4 , log_lik=TRUE )
```

Again, we get the summaries of the posterior distributions of the
parameters, and assemble their posterior means in a data frame:


```{r }
## Posterior summaries
(m2.cf <- precis(m2, depth = 2, digits=3))
## Data-frame with posterior means of slope and intercept
## for each site
m2.ab <- data.frame(
    site = 1:nsites,
    site_fac = levels(islands$site_fac),
    intercept = m2.cf[1:nsites,1],
    slope = m2.cf[(nsites+1):(2*nsites),1])
```

And them we can add the regression lines of this model and the of the
previous one to the scatter plots:


```{r }
p2 +
    geom_abline(data = m1.ab,
                 aes(intercept = intercept, slope = slope),
                col="navy") +
     geom_abline(data = m2.ab,
                 aes(intercept = intercept, slope = slope),
                col="red")
```

Pooled slopes (red lines) are less variable, because the fit to each
site uses information from the other sites. Thus, the parameters
shrink towards a common mean:

```{r }
plot(m1.ab$slope, rep(1,nrow(m1.ab)),
     ylim = c(0.5, 2.5),
     xlab = "Mean posterior slopes",
     yaxt="n", ylab="" , cex.axis=2,
     cex.lab = 2, pch = 19)
points(m2.ab$slope, rep(2,nrow(m2.ab)), pch =19)
segments(y0 = rep(1,nrow(m1.ab)), x0 = m1.ab$slope,
         y1 = rep(2,nrow(m1.ab)), x1 = m2.ab$slope,
         lty =2)
axis(2, at = c(1, 2), labels = c("M1", "M2"), cex.axis=2, adj = 1, las = 2)       
```

This common mean is the hyperparameter $\mu_\beta$. Add the Bayesian
confidence interval of this parameter to the plot above:


```{r }
segments(x0 = m2.cf["mu_b","5.5%"],
         y0 = 2.2,
         x1 = m2.cf["mu_b","94.5%"],
         y1 = 2.2, lwd=2, col ="blue")
```

This model assumes a normal distribution for the slopes of the
regression, which is the parameter of interest in an analysis of
species-area relationships. 

How variable would be these slopes among archipelagoes, if our model
provides a good prediction? We can assess this by simulating the
posterior distributions of the slopes. To do that, we first take 
samples from the posterior distributions of the mean and standard
deviations of the normal that describe the distribution of slopes. To
simulate an archipelago taken at random, we then simulate a sample,
from a normal using these parameters. By repeating this many times we
can approximate the posterior distribution of the slopes. For
reference, we added also the prior distributions of this slopes used
in the fit:

```{r }
## Sample posteriors of mean and standard deviation of slope distributions
m2.samp <- extract.samples(m2, pars = c("mu_b", "sigma_b"))
## Simulate normal samples using the parameter sampled above
sim.slopes <- rnorm(length(m2.samp$mu_b), mean = m2.samp$mu_b, sd = m2.samp$sigma_b)
## Simulate prior of slopes
m2.priors <- data.frame(mu_b = rlnorm(1e4, meanlog = 0, sdlog = 0.7),
                        sigma_b = rexp(1e4, rate =1))
sim.prior.slopes <- rnorm(length(m2.priors$mu_b), mean = m2.priors$mu_b, sd = m2.priors$sigma_b)
## Densities of the simulated prior and posterior
plot(density(sim.slopes),
     xlab = "Slopes", main ="",
     xlim = c(-2,5))
lines(density(sim.prior.slopes), col ="blue")
legend("topleft", legend = c("Prior", "Posterior"),
       lty =1, col=c("blue","black"), bty = "n")
```

### Model comparison

The package *rethinking* has the handy function
[compare](https://rdrr.io/github/rmcelreath/rethinking/man/compare.html)
that returns a model selection table using WAIC (Gelman *et al.*
2013). As for any information-based criteria, the model with best
predictive information (or with the smaller loss of information of the
data-generating process) has the smaller value of WAIC:

```{r }
compare(m0, m1, m2)
```

The WAIC tells us that the model with polled intercepts and slopes is
as good to predict future observations as the model with independent
slopes and intercepts. This suggests that the former may be overfitted.


## The maximum likelihood way

There are functions and libraries in R with efficient routines to fit
many classes of models with the maximum likelihood method. Most of these functions use
the [model formula
syntax](https://www.ucd.ie/ecomodel/Resources/R_formulae_WebVersion.html),
which expresses the structure of the linear (or non-linear) predictor.


### Single slope an intercept

This is the simple linear regression. As we [have seen already](https://statistical-inference.netlify.app/03_linear_models.html), we fit
this model by the ML method with the function`lm`:


```{r }
m0.ML <- lm(lc_S ~ ls_area, data = islands)
```

And here are the estimates and confidence intervals of the estimated
parameters of the linear predictors:

```{r }
## Intercept and slopes estimates and CI
coef(m0.ML)
confint(m0.ML)
```

The estimate of the residual standard deviation $\sigma$ is returned by


```{r }
sigma(m0.ML)
```

To get the limits of the frequentist 95% confidence interval for
$\widehat{\sigma}$ of this model we apply the expression:

$$
\left(\,  \sqrt{ \frac{\text{RSS}}{\chi^2_{0.975,n-2}}} \; , \; \sqrt{ \frac{\text{RSS}}{\chi^2_{0.025,n-2}}} \, \right)
$$

Where $\text{RSS}$ is the sum of squared residuals, $n$ is the number
of observations and $\chi^2_{p,n-2}$ is the quantile of the
Chi-square distribution for a probability of $p$ and $n-2$ degrees of
freedom^[in this case, the degrees of freedom of the sum of squared residuals is the number of observations minus the number of parameters of the linear predictor].

Here is how to calculate the CI for $\hat{\sigma}$ in R:


```{r }
RSS <- sum(residuals(m0.ML)^2)
N <- nrow(islands)
## Lower CI limit
sqrt( RSS / qchisq(0.975, df = N -2))
## Upper CI limit
sqrt( RSS / qchisq(0.025, df = N -2))
```

**Question:** do the parameter estimates and their confidence intervals of the ML and Bayesian fitting match?


### Different slopes and intercepts for each site

This is multiple linear regression, with island area and the site
identity as predictor variables. Site identity is a categorical factor
with as many levels as sites:


```{r }
## site id with sites names as labels
levels(islands$site_fac)
## site id with numeric labels (the same)
levels(islands$site_id)
```

By including an additive effect of site id in the model you allow for
different intercepts^[that is, the mean log richness differ among
sites]. And by including an interaction between island area and site
id you allow for different slopes for each site^[that is, the effect
of island area varies between sites]:


```{r }
m1.ML <- lm(lc_S ~ ls_area + site_id + ls_area:site_id, data=islands)
```

And then you get the estimated coefficients with:

```{r }
(m1.ML.cf <- coef(m1.ML))
```

The default structure of the linear predictor for a multiple
regression with categorical variables like that in R is

$$mu_i = \alpha_0 + \alpha_{site[i]} + (\beta_0 + \beta_{site[i]}) \, \text{ls_area}$$

Where $\alpha_0$ is the intercept for the first level of the
categorical variable (the site coded as #1), $\beta_0$ is the slope
also for the first level of the categorical variable. And thus
$\alpha_{site[i]}$ are the **differences between intercepts** for
observations at the level 1 of the categorical predictor and level
$i$. In the same way, $\beta_{site[i]}$ are the differences between
the slopes for level 1 and level $i$.

Therefore, in our model the coefficients labeled as *"(Intercept)"*
and *"ls_area"* are the intercept and slopes estimated for the study
site 1. You can get the estimates of intercepts for sites 2 to `r nsites` with:


```{r }
m1.ML.cf[1] + m1.ML.cf[3:(nsites+1)] 
```

**Question:** Calculate the estimated slopes for each site and compare
with the corresponding estimates by the Bayesian fit. Do they match?
After making your comparisons, click below to unfold a code to draw
the regression lines from the Bayesian (blue) and Ml (red) fittings.

```{r code_folding=TRUE}
## Assembles a data-frame with intercepts and slopes
## for the ML fitting
m1.ML.ab <- data.frame(
    site = 1:nsites,
    site_fac = levels(islands$site_fac),
    intercept = NA,
    slope = NA)
m1.ML.ab$intercept[1] <- m1.ML.cf[1]
m1.ML.ab$slope[1] <- m1.ML.cf[2]
m1.ML.ab$intercept[2:nsites] <- m1.ML.cf[3:(nsites+1)]+m1.ML.cf[1]
m1.ML.ab$slope[2:nsites] <- m1.ML.cf[(nsites+2):(2*nsites)]+m1.ML.cf[2]
## Adds the lines for both fitting methods to the scatter-plot
p2 +
    geom_abline(data = m1.ab,
                aes(intercept = intercept, slope = slope),
                col="navy") +
    geom_abline(data = m1.ML.ab,
                aes(intercept = intercept, slope = slope),
                col="red")
```

### Pooled intercepts and slopes

In the approach of ML methods, this is a **mixed effects** linear model, which is specified as:

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma) \\
  \mu_i &= \alpha + a_{\text{site[i]}} + (\beta + b_{\text{site[i]}}) \, x_i \\
  a_{\text{site}} & \sim \text{Normal}(0, \sigma_a)\\
  b_{\text{site}} & \sim \text{Normal}(0, \sigma_b)
\end{aligned}
$$

Where $\alpha$ and $\beta$ are called **fixed effects** and express
the average intercept and slope for all observations. There is a
parameter $a_{\text{site}}$ for each site, that follows a normal
distribution with zero mean. These are the **random effects** of site
on the intercept, that are centered at the value of $\alpha$, and
change the fixed intercept by a given amount, for each
site. Accordingly, the parameters $b_{\text{site}}$ are the random
effects of site on the slopes. 

The reference package to fit mixed-effects models by the ML methods is
[lme4](http://cran.nexr.com/web/packages/lme4/vignettes/lmer.pdf). The
function `lmer` fits Gaussian linear mixed models. To fit the model
with random effects on intercepts and slopes use:


```{r }
m2.ML <- lmer(lc_S ~ ls_area + (ls_area|site_fac), data = islands, REML=FALSE)
```

In the syntax of `lmer`, the term `(ls_area|site_fac)` means that site
is a random effect for the fixed linear predictor on `ls_area`. This
term thus adds random effects to the intercept and to the slope of the
linear predictor^[actually the term *(ls_area|site_fac)*  is a
shorthand for *(1|site_fac) + (ls_area|site_fac)*, which explicitly
indicates a random effect for the intercept and other random effect
for the slope. Thus a model with random effects only in the slope
would include only the term *(1|site_fac)*].

A summary of the model shows the estimates of the variance of the
random effects (parameters $\sigma_a$ and $\sigma_b$), and the
estimates for the fixed effects (parameters $\alpha$ and $\beta$),
and some more important figures:

```{r }
summary(m2.ML)
```

**Question:** compare the estimated standard deviations of the random
effects with those estimated by the Bayesian fit of the corresponding
model. Do they match?  **Hint:** The call `precis(m2)` will show you a
summary of the posteriors that you need.


The estimates for the coefficients for each site are returned by:


```{r }
(m2.ML.cf <- coef(m2.ML)$site_fac)
```

We can easily add to this data frame the site names and then use this
to compare the regression lines for each site fitted by ML (green) and
by the Bayesian (red) way:


```{r }
## Adds a factor with the name of sites
m2.ML.cf$site_fac <- factor(rownames(m2.ML.cf))
## Change columns names to fit to the ggplot call
names(m2.ML.cf)[1:2] <- c("intercept", "slope")
## Call the plot and add lines
p2 +
    geom_abline(data = m2.ML.cf,
                 aes(intercept = intercept, slope = slope),
                col="green") +
     geom_abline(data = m2.ab,
                 aes(intercept = intercept, slope = slope),
                col="red")
```

**Questions:** why the lines do not match exactly? Why some match more than others?

### Model comparison

Use the `bbmle::AICctab` function to build a model selection table using AIC corrected for small samples:


```{r }
AICctab(m0.ML, m1.ML, m2.ML, nobs=nrow(islands))
```
According to this model selection, fitting independent intercepts and slopes is definitely overfitting.


# References and further reading

* Bates, D., Mächler M., Bolker, B.M. and Walker, S.C.. 2015. Fitting Linear Mixed-Effects Models Using lme4. Journal of Statistical Software 67 (1): 1-48.
* Bolker, B. M. et al. 2009. Generalized linear mixed models: a practical guide for ecology and evolution. Trends in ecology & evolution, 24(3): 127-135.
* The Generalized Linear Mixed Models FAQ page: https://bbolker.github.io/mixedmodels-misc/glmmFAQ
* Gelman, A., Hwang, J., & Vehtari, A. (2014). Understanding predictive information criteria for Bayesian models. Statistics and computing, 24(6), 997-1016.
* Gelman, Andrew, and Jennifer Hill. 2006. Data Analysis Using Regression and Multilevel/Hierarchical Models. Cambridge, England: Cambridge University Press.
* Ohyama, L., Holt, R. D., Matthews, T. J., & Lucky, A. (2021). The
  species–area relationship in ant ecology. Journal of Biogeography,
  48(8), 1824-1841.
* Zuur, A et al. 2009. Mixed Effects Models and Extensions in Ecology with R. Springer.
