---
title: "Multiple linear regression Lab"
author:
 - "Andrea Sánchez-Tapia"
 - "Paulo Inácio Prado"
date: "2024-02-01"
output:
    distill::distill_article:
      toc: true
      toc_depth: 3
---

```{r setup, echo=FALSE}
library(rmarkdown)
knitr::opts_chunk$set(eval = FALSE, fig.asp = 1)
# clipboard
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "Copy code <i class=\"fa fa-clipboard\"></i>",
    success_text = "Copied! <i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "Not copied 😕 <i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
  )

```


# Introduction 

We can extend the simple linear regression model to accommodate
multiple predictors. These are the **multiple linear regression**
models, of which the simple linear regression is just a particular
case with a single predictor variable.


In the next sections we will show how to fit multiple regression
models with the least square methods in R, and how to interpret the
model coefficients.

# Additive multiple linear model

In an additive multiple linear regression, the expected value of the
response $\mu$ is the sum of multiple predictor variables
$X_i$, each one weighted by its own coefficient
$\beta_i$ :


$$
\begin{align}
y_i &\sim Normal(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_j x_{ij} \\
\sigma & = C
\end{align}
$$

The expression for the predicted value in this model can be written in the compact form:

$$\mu_i = \sum_{j=0}^{J} \beta_{j} x_{ij}$$

Where $J$ is the number of predictor variables. This is called the **linear predictor**.

## Fitting

We will use a data set of an experiment done with the biennial plant
*Ipomopsis aggregata*^[Crawley 2007. The R Book, Wiley]. Forty plants
were assigned at random to two treatments: unprotected from grazing by
rabbits and protect from grazing by fences. The fruit yield of each
plant (dry mass in mg) was measured at the end of the experiment. To
take into account the effect of the size of the plant on fruit
production, the diameter of the top of rootstock (mm) was also
recorded. Copy the following commands and paste in your R console to
read and prepare this data for the analyses:

```{r data, eval=TRUE}
## Data reading and preparation
ipomopsis <- read.csv("https://raw.githubusercontent.com/lmmx/crawley-r-statistics-book-notes/master/code/ipomopsis.csv")
## More proper names
names(ipomopsis) <-  c("diameter","fruit.w","treatment")
## Convert treatment to a factor variable
ipomopsis$treatment <- factor(ipomopsis$treatment,
                              labels = c("Control", "Fenced"))
```

Let's start with a plot of the response variable as function of both
predictors:


```{r plot}
plot(fruit.w ~ diameter, data = ipomopsis, type ="n",
     ylab = "Fruit dry weight (mg)",
     xlab = "Rootstock diameter (mm)")
points(fruit.w ~ diameter, data = ipomopsis,
       subset = treatment == "Fenced")
points(fruit.w ~ diameter, data = ipomopsis,
       subset = treatment == "Control", pch=2, col ="red")
legend("topleft", c("Fenced", "Control"), pch=c(1,2),
       col =c("black", "red"), bty="n")
```

Larger plants tend to produce larger dry biomass of fruits. It also
seems that if we control for size, non-fenced plants had larger fruit
yields. We can check this hypothesis by fitting a linear regression
with the additive effect of both plant size and treatment:


```{r  lm, eval=TRUE}
ipo_m1 <- lm(fruit.w ~ diameter + treatment, data = ipomopsis)
```

Use the function `coef` to store the coefficients ($\beta_i$) of this
fitting in a object:


```{r coef, eval=TRUE, fig.asp = T}
(ipo_cf1 <- coef(ipo_m1))
```

The predictor variable "treatment" is a factor with two levels
(control/fenced). The default in R is to create dummy variables for
each level of the factor, except the first one in alphabetical
order. This means that the model has a variable *"treatmentFenced"*,
that has value one for fenced plants, and value zero otherwise. So,
fruit yield predicted ($\widehat{Y}$) by this model for control plants is:

$$\widehat{Y}_{\text{fenced}} = \beta_0 + \beta_1 \text{diameter} + \beta_2 \times 0$$

While for fenced plants we have:

$$\widehat{Y}_{\text{non-fenced}} = \beta_0 + \beta_1 \text{diameter} + \beta_2 \times 1$$


To add the line of expected yield for fenced plants we can use the
function `abline`, that draws a straight line with a given intercept and slope:

```{r plot2}
## Make the plot again
plot(fruit.w ~ diameter, data = ipomopsis, type ="n",
     ylab = "Fruit dry weight (mg)",
     xlab = "Rootstock diameter (mm)")
points(fruit.w ~ diameter, data = ipomopsis,
       subset = treatment == "Fenced")
points(fruit.w ~ diameter, data = ipomopsis,
       subset = treatment == "Control", pch=2, col ="red")
legend("topleft", c("Fenced", "Control"), pch=c(1,2),
       col =c("black", "red"), bty="n")
## Adds the predicted line for fenced plants
## check the function help to use it
abline( a = ipo_cf1[1], b = ipo_cf1[2], col = "red")
```

**Challenge:** now add to the plot a black line for the expected yield for
non-fenced plants. Your plot should look like this:


```{r plot3, echo = FALSE, eval =TRUE}
## Make the plot again
plot(fruit.w ~ diameter, data = ipomopsis, type ="n",
     ylab = "Fruit dry weight (mg)",
     xlab = "Rootstock diameter (mm)")
points(fruit.w ~ diameter, data = ipomopsis,
       subset = treatment == "Fenced")
points(fruit.w ~ diameter, data = ipomopsis,
       subset = treatment == "Control", pch=2, col ="red")
legend("topleft", c("Fenced", "Control"), pch=c(1,2),
       col =c("black", "red"), bty="n")
## Adds the predicted line for fenced plants
## check the function help to use it
abline( a = ipo_cf1[1], b = ipo_cf1[2], col = "red")
abline( a = ipo_cf1[1]+ipo_cf1[3], b = ipo_cf1[2])
```

From the plot above, we can see that the difference between the
yield of plants of the same size is a constant, that corresponds to
the value of the coefficient of the variable *"treatmentNon-Fenced"*.
Thus, the fitted model estimates that non-fenced plants had on average
more `r round(ipo_cf1[3])` mg of fruit yield, if we control for plant
size.

**Final Challenge** 

What if we had not controlled for plant size? Fit a
model only with the treatment and interpret the coefficients. 

Hint: it
is good practice to do exploratory data analyses before any model
fitting. In this case a box plot can help, try: 

```{r boxplot}
boxplot(fruit.w ~ treatment, data=ipomopsis)
```


# Adding interactions to the model


In this example, beds of tulips grown in greenhouses are subjected to
different soil moisture and light conditions. The response variable is
the size of blooms under these varying conditions.



```{r tulips, eval = T}
library(rethinking)
data(tulips)
str(tulips)
```

The variable `water` indicates one of three ordered levels of soil
moisture, from low (1) to high (3). The variable `shade` indicates one
of three ordered levels of light exposure, from high (1) to low (3).

Both predictor factors are crucial for plant growth. In simple
regression settings, we would expect a positive response of bloom size
to increases in soil moisture and light (in this case a negative slope
for `shade`).

However, how these two variables interact to affect plant growth is
also interesting. Higher soil moisture in the absence of light limits
photosynthesis, that requires both. Likewise, in the absence of water,
higher luminosity does not contribute to plant growth.

The mutual dependency of how predictor variables affect the response can be modeled as an
__interaction__.

The model with interaction has the following notation: 

$$
\begin{align}
y_i & \sim Normal(\mu_i, \sigma) \\
\mu_i & = \beta_0 \, + \, \beta_1 x_{i1} \, + \, \beta_2 x_{i2} \, + \,  \beta_3 x_{i1}  x_{i2}\\
\sigma & =  C
\end{align}
$$ 

Where:

$y_i$ is each value of the response variable *blooms*

$x_{i1}$ is each value of the predictor variable *water*

$x_{i2}$ is each value of the predictor variable *shade*

$x_{i1} \dot x_{i2}$ is the product of the two explanatory variables, or the **interaction term**.

$\beta_j$ are the model coefficients.


Thus, model with interaction adds a third parameter $\beta_3$ to account
for the interaction between the level of watering and shading.


```{r center, echo = F, eval = F}
#Let's center the data:
blooms <- tulips$blooms / max(tulips$blooms) 
water <- tulips$water - mean(tulips$water) 
shade <- tulips$shade - mean(tulips$shade)
```

Let's plot the data: 

```{r plot_tulip}
color <- c("black", "red", "blue")
plot(blooms ~ water, 
     data = tulips, 
     type = "n",
     ylab = "Bloom numbers",
     xlab = "Water level")
for (s in 1:3) {
points(blooms ~ water, 
       data = tulips,
       subset = shade == s, pch = s, col = color[s])
}
legend("topleft",
       c("Shade 1", "Shade 2", "Shade 3"),
       pch = c(1, 2, 3), 
       col =c("black", "red", "blue"), bty = "n")

```



```{r plot_three, fig.asp = 0.6, echo = F, eval = F}
#Let's plot three panels for this dataset: 
color <- c("black", "red", "blue")
par(mfrow = c(1, 3))
 #just to plot
for (s in 1:3) {
  index <- which(tulips$shade == s)
 plot(blooms ~ water, data = tulips[index,], type = "n",
     ylab = "Bloom numbers",
     xlab = "Water level")
points(blooms ~ water, data = tulips[index,],
       subset = shade == s,
       pch = s, col = color[s])
title(paste("Shade", s))
}

```


Let's fit the model with the interaction between the effects of shade
and water, and extract its coefficients. The R syntax to include an
interaction in a model formula is *"variable1:variable2"*.

```{r interactions}
tulip_int <- lm(blooms ~ water + shade +
                  water:shade, data = tulips)
(tul_cf2 <- coef(tulip_int))
```

Unlike the additive model, the slopes of the regression model with the
interaction factor can vary between groups.

In this example, the interacting (or moderator) variable `shade` has
only three values (1, 2, 3).  We could treat both variable as factors
with three levels, but for learning purporses we will use it here as a
continuous value. Thus the predicted values for the lower levels of
water (*water = 1*) for any value of shade is:

$$\begin{align}
\widehat{Y}_{[\text{water=1}]} & = \beta_0 \; + \; \beta_1 \times 1 \; + \; \beta_2 \times \text{shade} \; +  \; \beta_3 \times 1 \times \text{shade} \\
& = (\beta_0 + \beta_1) \; + \; (\beta_2 + \beta_3) \times \text{shade}
\end{align}
$$

That is, if the value of the variable is one, the intercept of the
linear predictor is increased in $\beta_1$, and the slope of is
increased in $\beta_3$.

The predicted values for *water = 2* are:

$$\begin{align}
\widehat{Y}_{[\text{water=2}]} & = \beta_0 \; + \; \beta_1 \times 2 \; + \; \beta_2 \times \text{shade} \; +  \; \beta_3 \times 2 \times \text{shade} \\
& = (\beta_0 + 2 \beta_1) \; + \; (\beta_2 + 2 \beta_3) \times \text{shade}
\end{align}
$$

And so on. Therefore, in a model with interaction, the *effect* of one
predictor variable (that is, how much its increase affects the
response), depends of the value of some other predictor variable. In
this example, the effect of increasing or decreasing shading depends
on how moits are the plant beds.

Let's calculate conditional slopes for the different values of shade
(1, 2, and 3). A handy way to do that is:

```{r calculations}
s <- 1:3 #values of shade
intercepts <- tul_cf2[1] + tul_cf2[3] * s
slopes <- tul_cf2[2] + tul_cf2[4] * s
```

And now we can plot again the dataset and add lines according to these values. 

```{r plotagain}
color <- c("black", "red", "blue")
plot(blooms ~ water, 
     data = tulips, 
     type = "n",
     ylab = "Bloom numbers",
     xlab = "Water level")
for (s in 1:3) {
points(blooms ~ water, 
       data = tulips,
       subset = shade == s, pch = s, col = color[s])
  abline(a = intercepts[s], slopes[s], col = color[s])
}
legend("topleft",
       c("Shade 1", "Shade 2", "Shade 3"),
       pch = c(1, 2, 3), 
       col = c("black", "red", "blue"), bty = "n")

```

As you can see, the effect of more soil humidity is steep when there
is more light, and it decreases with increasing values of shade.

__Extra:__ 

+ Run the diagnostic plots for this model 

+ Fit and compare the model with only additive effects for this dataset:

$$
\begin{align}
y_i & \sim Normal(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}\\
\sigma & =  C
\end{align}
$$ 

+ Fit the model after scalling the predictors so they are centered at zero:

```{r}
tulips_centered = within(tulips,{
  water = water - mean(water)
  shade = shade - mean(shade)})
```

What happens with the coefficients? What happens with the predicted values?

# References

+ McElreath, R., 2018. Statistical Rethinking: A Bayesian Course with Examples in R and Stan, 2nd ed. CRC Press.


