---
title: "Random variables and probability distributions"
subtitle: "Serrapilheira/ICTP-SAIFR Training Program in Quantitative Biology and Ecology - 2023"  
author: 
  - "Andrea Sánchez-Tapia"
date: 'January 10th-11th 2023'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
self-contained: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(servr.daemon = TRUE)#para que no bloquee la sesión
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#03045E",
  colors = c(
    red = "#A70000",
    white = "#FFFFFF",
    black = "#181818"
  ),
  text_bold_color = "#03045E",
  header_font_google = google_font("Roboto Condensed"),
  text_font_google = google_font("Roboto Condensed", "300", "300i"),
  code_font_google = google_font("Fira Mono"), text_font_size = "32px",
  code_font_size = "32px"
)
xaringanExtra::use_share_again()

# clipboard
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "Copy code <i class=\"fa fa-clipboard\"></i>",
    success_text = "Copied! <i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "Not copied 😕 <i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)

```

 

## Today's key concepts

+ Random variables
+ Discrete random variables: Bernoulli, Binomial, Poisson.
+ Continuous random variables
    + Probability density functions, PDFs
    + Cumulative density functions, CDFs
+ Expected value, variance
+ Gaussian distribution and its key properties
+ Central limit theorem and the Law of large Numbers

---

## Empirical data

+ Assuming __random__ and __independent__ sampling 

--

+ We study variables for which __the specific outcome of a single trial or measurement is unknown__ even if we understand the set of all possible outcomes (the sample space)

--

+ Given enough experiments/trials, we can hypothesize/estimate how often the possible outcomes will happen: their __probability distribution__ 

--
    + A coin toss/the presence of a species
--
    + The number of individuals in a given sample
--
    + The length of a body part

---
## Random variables

+ __Random variables__ are mathematical formalizations of these several types of data

--

+ Some experiment outcomes are __discrete__, countable and some are not, they can take any value inside __continuous__ intervals.

--

+ __Probability distributions__: they represent __theoretical__ distribution of a random variable of a certain kind. (How probable are each of the outcomes)

--

+ Random variables can be described by their __parameters__: key properties that allow us to reconstruct their probability distributions

???
somebody talked yesterday about "probabilistic models"

---
class: inverse, middle, center 

# Discrete random variables

---
## Bernoulli distribution

+ The simplest experiment has only two outcomes. A coin flip. 

--

+ The probability $p$ and $q$ of the two outcomes sums 1. $p + q = 1$   

--

+ If equiprobable, then $p = q = 0.5$ 
$$X \thicksim  Bernoulli(p)$$
--

+ Repeated Bernoulli trials form a __Binomial Random Variable__. The parameter $n$ = number of trials is key to understanding Binomial Random Variables


---
## Binomial distribution 

$$X \thicksim  Bin(n,p)$$

--

+ The number of successes $X$ in $n$ independent Bernoulli trials with $P(success) = p$

--

+ The probability of $X$ successes: $p^X$

--

+ The probability of $n-X$ failures: $(1-p)^{(n-X)}$

--

+ The number of __combinations__ of X successes in n trials: $$\binom{n}{X}$$

---
class: inverse 

What is the probability of 11 successes in 25 trials when $p = 0.5$?

```{r}
n <- 25
p <- 0.5
X <- 11
P_success <- p^X
P_failure <- (1 - p)^(n - X)
combinations <- choose(n, X)
P_success * P_failure * combinations
dbinom(X, size = n, prob = p)
```

---
### We can calculate P(X) for every X in 0:25

```{r, eval = F}
P_X <- dbinom(1:25, size = n, prob = p)

plot(P_X, xlab = "X")
title(
  "Probability distribution of X ~ Binomial(25, 0.5)")
abline(v = X ,col = "red")
abline(h = dbinom(X, size = n, prob = p) ,col = "red")
```

---
### We can calculate P(X) for every X in 0:25

```{r, echo = F}
P_X <- dbinom(1:25, size = n, prob = p)
plot(P_X, xlab = "X")
title("Probability distribution of X ~ Binomial(25, 0.5)")
abline(v = X ,col = "red")
abline(h = dbinom(X, size = n, prob = p) ,col = "red")
```


<!--
```
dbinom(x, size, prob)
pbinom(x, size, prob)
qbinom(p, size, prob)
rbinom(n, size, prob)
```

+ `size` is $n$, the number of trials 
+ `prob` is $p$, the probability of a success in a single Bernoulli trial
+ `dbinom` gives the probability density for the values in `x`
+ `pbinom` gives the cumulative density of the distribution (e.g. for values equal or lesser than `x`)
+ `qbinom` gives the value in x that correspond to the cumulative probability p
+ `rbinom` samples n values from the distribution defined by $n$ (`size`) and $p$ (`prob`)
-->


---

## Poisson random variables

 <!-- + n large, p small, ex. rare plants, animals, unobserved trials --> 

+ The number of events (__counts__) in samples of fixed area, or during a fixed interval of time

--

+ Each sample independent from the other

--

+ The starting point of such a distribution is zero, a count. 

--


+ A single parameter: $\lambda$, a _rate_ parameter: the __average number of events per sample__. 

--

$$X \sim Poisson(\lambda)$$

---
## Poisson random variables

+ The probability of any observation is: 

$$P(X) = \frac{\lambda^X}{X!} e^{-\lambda} $$  

--

+ When the event is not very frequent, a reverse-J-shaped curve with a long right tail 

--

+ When the event becomes more frequent, the most frequent value increases too and the curve becomes more symmetrical.

--

+ With large lambda values, the distribution approaches the shape of a normal curve (but remember it is discrete!)


 <!-- Relationship between binomial, Poisson and normal distributions
 A binomial can be approached to a Poisson when n is large compared to p
 When \lambda increases, the Poisson distribution becomes more symmetrical and approaches a normal distribution. 
 Binomial between 0 and n (number of trials)
 Poisson is not bounded, long left tail --> 

---
class: inverse

```{r, echo = TRUE, eval = TRUE}
p_poisson <- function(x, lambda) {
  (lambda ^ x / factorial(x)) * exp(-lambda)
}
P_x <- p_poisson(0:5, lambda = 1)
#P_x <- dpois(0:5, lambda = 1)
P_x
#plot(P_x, xlab = X)

```

---

## Expected value of discrete variables

+ __Weighted average__, where each value is given the importance according to its probability 

--

$$ E(X) = \sum_{1=1}^{n}{a_ip_i}$$

--

+ Bernoulli: $p$
+ Binomial: $np$
+ Poisson: $\lambda$

--

Important: This expectation just marks the tendency and may be outside of the range of possible values of $X$. 

---
## Variance of a discrete random variable

$$\sigma^2(X) = E[X-E(X)]^2$$
--

+ Bernoulli: $p(1-p)$
+ binomial: $np(1-p)$
+ Poisson: $\lambda$

---
class: inverse, center, middle

# Continuous random variables


---

## Continuous random variables

+ Measurements (length, biomass, weight) in a continuous scale, imposible to count or to define the specific set of outcomes

--

+ Precision depends on measurement

--

+ The random variable is represented by a __probability density function__, a continuous curve depicting "relative" probabilities

---
## Probability density function (PDF)


+ Total area under the PDF curve equals 1

--

+ The specific probability of a single value is zero. Asking this does not make sense because there are infinite possible values
    
--

+ Instead, We define the probability of $X$ belonging to an interval.

--

+ The probability is __the area under the curve__ within the interval: the __integral__

---

## Cumulative distribution function (CDF)

+ The probability of $X$ being lower than $Y$
$F(Y) = P(X < Y)$ is the area under the PDF curve in the interval where $X < Y$

+ The CDF sums the areas below Y and therefore is an ever-increasing (monotonic) curve from 0 to 1


---
## Uniform random variable

$X \thicksim {Uniform[a,b]}$

+ Defined for an interval $[a, b]$

--

+ The probability of sub-intervals of the same width is the same $[a, b]$

--

+ The Distribution Probability Function is $1/b-a$ for all the $[a, b]$ interval. (The area under the curve is "divided" by the width of this interval)


<!-- $E(X) = \frac{b + a}{2}$
$\sigma^2(X) = \frac{(b - a)^2}{12}$-->



---

## Expected value 

+ For any particular observation, 0
+ Subintervals $\Delta x$

$$\sum_{i=1}^{n}{x_if(x_i)\Delta x}$$
$$E(X) = \int{xf(x)dx}$$

---
## Normal random variables

+ "Gaussian" or normal distribution

--

+ The basis for linear regression and analysis of variance

--

+ Fits many empirical datasets

--

+ Measurements, symmetric around some average value, with some variation around this value

--

$$X \thicksim{\mathcal{N}(\mu, \sigma})$$
--

+ $E(X) = \mu$, Variance: $\sigma^2(X) = \sigma^2$

--

$$\mathcal{N}(\mu, \sigma) = \frac{1}{\sigma \sqrt{2\pi}}\;e^{-\frac{1}{2}\left(\frac{Y-\mu}{\sigma}\right)^2} $$

---
## Normal random variables

+ Standard normal random variable: $Z$ $\mu = 0$ and $\sigma = 1$

--

+  Standardizing: substract the mean and divide by the standard deviation

--

+ Any random variable can be standardized that way.

---
## Other continuous random variables

+ __Log-normal__ 
    + $X$ such that $ln(X)$ is a normal random variable. 
    + Skewed, long tail to the right 
    
    $$E(X) = e^ {\frac{2\mu + \sigma^2}{2}}$$

+ __Exponential__
    + Related to Poisson. Rare events in a continuous space. 
+ Others... Student-t, $\chi^2$, F, gamma, inverse gamma, beta.

---
## The Central Limit Theorem

+ Several _independent_ random variables with the same expected value and variance, any distribution 

--

- Their sum or average can be standardized to a standard normal random variable ( $Z$ ) _provided the sample size is large enough_.

--

- "Arithmetic means of large samples or random variables conform to normal distribution even if the underlying distribution does not"


### This allows using statistical tests that assume/require normal distributions. 

---
## Law of Large Numbers 

Given increasing sampling size, the means of samples will converge towards the means of the population, $\mu$. 

---
## References

+ Gotelli, N.J., Ellison, A.M., Ellison, S.E. and S.R.F.H.F.A.M., 2004. A Primer of Ecological Statistics. Sinauer Associates Publishers.

+ McElreath, R., 2018. Statistical Rethinking: A Bayesian Course with Examples in R and Stan, 2nd ed. CRC Press.
