---
title: "Data simulation for linear models"
author:
 - "Andrea Sánchez-Tapia"
 - "Paulinha Lemos-Costa"
date: "2023-01-13"
output:
    distill::distill_article:
      toc: true
      toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Simulating data according to linear models


Let's simulate the relationship of Weight as a function of Height in Humans

$$Height \rightarrow Weight $$

+ We know that $\beta_0$ is the intercept. If $\beta_0 = 0$ this means that an individual with no height has also no weight, which makes sense...

+ We can also imagine that for every 10 centimeters a person grows, their weight increases 5kg. (It's just an example). So the slope of our linear relationship would be $\beta_1 = 0.5$

$$
W_i \sim \mathcal{N}(\mu_i, \sigma) \\
\mu_i = \beta_0 + \beta_1 X_i
$$

```{r height}
beta_0 <- 0
beta_1 <- 0.5
sigma <- 5
n_individuals <- 100

H <- runif(n_individuals, 130, 170) # Why a uniform distribution? remember H is our predictor variable! we are sampling numbers in the interval of heights for human adults, but we don't want any particular value to be more probable in this interval.

#According to our linear model:
mu <- beta_0 + beta_1 * H
# The response variable W is therefore distributed as a normal random variable with mean = mu and sigma = sigma
W <- rnorm(n_individuals, mu, sigma)

# If we plot our synthetic data: 
plot(W ~ H, xlab = "Height (cm)", ylab = "Weight (kg)")

```



With this data, let's fit a linear regression: 

```{r}
humans <- data.frame(Weight = W, Height = H)
sim_model <- lm(W ~ H) 
coef(sim_model)
```

The coefficients of this simulated dataset are not exactly the same parameters that we defined above during the simulation. This makes sense since they come form a sample from the whole distribution.

 <!--we can rescale later with  X-mean(X) so the intercept  means having the average weight--> 
 
 
